### 2016-06-13 Projects/DelCo_2014_2015/Reports/Hose_Stream_Tests/TeX_Files/DelCo_Hose_Stream_Tests.tex
Updated hose stream report with Dan's edits

%http://www.tft.com/literature/library/files/bfa.pdf
%Keith Royer, "Water for Fire Fighting", (Engineering Extension Service, Iowa State University, Ames, Iowa) p. 6 
In addition to the type of stream pattern used, the method used to apply the hose stream can also vary. For example, Royer suggested using a clockwise rotation instead of a counterclockwise application pattern for the following reasons~\cite{Royer:ISU}:
\begin{enumerate} 
	\item Clockwise rotation drives most of the heated gases, smoke, and flames away from the nozzle, while counterclockwise rotation does just the opposite.
	\item Steam formed by the clockwise rotation has a violent rolling action. Counterclockwise rotation produces steam with an inactive and lazy action.
	\item A clockwise rotation increases the efficiency of water and produces a faster knockdown time than counterclockwise rotation.
\end{enumerate}

\section{Objectives}
\label{sec:objectives}
This report is concerned with examining the impact of hose stream and application pattern, or nozzle motion, selection on air movement inside two full scale test structures built to represent residential sized structures. Numerous experiments containing different flow path configurations were conducted to study the impact of hose stream and water application selection on ventilation in a structure. More specifically, the experiments were conducted to fulfill the following objectives:
\begin{enumerate}
	\item Determine which hose stream patterns and application patterns cause significant air movement in a structure 
	\item Determine the amount of air movement induced by different hose streams and application patterns

\end{enumerate}
Four types of hose streams and four different water application patterns were studied throughout the experiments described in this report. Three types of hose stream patterns using a combination nozzle were studied: straight stream, narrow fog stream, and wide fog stream. Water was flowed into the structure by having the hose in a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Additionally, a set of experiments comparing the impact of a straight stream from a combination nozzle and a solid stream from a smooth bore nozzle equipped with a 1~in tip on air flows in a structure was conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

@@

% =========================
% = EXPERIMENTAL OVERVIEW =
% =========================
\chapter{Experimental Overview}
\label{chap:exp_overview}
To study the impact of different hose stream and application pattern selections on air movement inside a structure, various water flow experiments were conducted in two experimental structures designed to replicate typical residential structures. The experiments focused on studying the differences in the amount of air movement caused by specific combinations of four different hose stream patterns and four different water application patterns.

@@

A total of seven experiments were conducted to study the impact of different hose stream and application patterns on gas movement in a structure; three used a combination nozzle attached to a monitor to flow water at approximately 120~gpm, three used a combination nozzle attached directly to a 1.75~in handline to flow water at approximately 120~gpm (Fig.~\ref{fig:monitor+handline}), and one used both a combination nozzle and a smooth bore nozzle attached to a monitor to flow water at approximately 180~gpm. Experiments that used a monitor to flow water always applied the stream in a fixed position and were focused on changing the location of water application in addition to the stream pattern, while the experiments that used a combination nozzle attached directly to a 1.75~in handline to flow water focused on varying the pattern in which water was applied in addition to the stream pattern.

@@

Three different structure layouts were used for the seven experiments. Tests~1 and 5 were conducted in the Single Story Structure and used the layout shown in Fig.~\ref{fig:east_setup}. Test~1 used a monitor and combination nozzle to flow water, and Test~5 used a combination nozzle attached directly to a handline to flow water.

Two different layouts were used for experiments conducted in the Two Story Structure. The main difference between the two layouts was the position of the south side door on the first level. Tests~2 and 6 used a configuration with the door opened (Fig.~\ref{fig:flow_path_1}), while Tests~3, 4, and 7 used a configuration with the door closed (Fig.~\ref{fig:flow_path_2}). Tests~2 and 3 used a combination nozzle attached to a monitor to flow water; Tests~6 and 7 used a combination nozzle attached directly to a handline to flow water; and Test~4 used both a combination nozzle and a 1~in tip smooth bore nozzle attached to a monitor to flow water in a straight and solid stream, respectively.

@@

\subsubsection{Test~1}
Test~1 utilized a monitor equipped with a combination nozzle to flow water from Room~C of the Single Story Structure to the area above the doorway on the south wall of Room~B (Fig.~\ref{fig:test_1_pic}). A plan view of the structure's layout during Test~1 can be found in Fig.~\ref{fig:east_setup}. The figure highlights the general direction of the established flow path along with the approximate location of A6, the bi-directional probe array used to measure gas velocity through the northeast interior doorway. The procedure for Test~1 was the simplest of all the test procedures. First, water was flowed in a straight stream for 30 seconds, then the stream was changed to a narrow fog for 30 seconds, and finally the stream was changed to a wide fog for 30 seconds. The procedure was repeated an additional time.

@@

\subsubsection{Tests~2 \& 3}
Tests~2 and 3 followed identical procedures that involved using a monitor equipped with a combination nozzle to flow water from the north side double doors to the interior ceiling on the first floor of the Two Story Structure. The nozzle was aimed at two targets during the series: a ``near'' target located approximately 1.8~m (6~ft) from the interior side of the north wall and a ``far'' target located approximately 3.7~m (12~ft) from the interior side of the north wall. The procedure for each test began by flowing water in a straight stream pattern at the far target for 60 seconds. After 60 seconds of water flow, the stairwell door was opened. One minute later, the west double door on the north side of the second floor was opened, and water continued to flow for 60 seconds. Next, the water flow was stopped, the two doors were closed, and the procedure was repeated with the monitor aimed at the near target. This entire process was repeated using the narrow fog and wide fog streams. Fig.~\ref{fig:test_2_3_pic} contains images of each type of hose stream aimed at the near and far targets.

@@

\subsection{Handline Experiments}
\label{sec:handline_procedure}
Three experiments were conducted using a 1.75~in handline with a combination nozzle to flow water in a straight stream, narrow fog stream, and wide fog stream. One test, Test~5, occurred in the Single Story Structure, and the remaining tests, Tests~6 and 7, occurred in the Two Story Structure. All three tests primarily focused on the impact of different hose stream and water application pattern selections on air movement in a structure.

@@

\subsubsection{Tests~6 \& 7}
Tests~6 and 7 followed identical procedures that involved using a 1.75~in handline equipped with a combination nozzle to flow water from the north side double doors of the Two Story Structure into the first floor room. The water stream was aimed at the south side wall of the first floor during the experiments. Test~6 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_1}, and Test~7 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_2}. Each experiment began by using a straight stream to flow water in a fixed, horizontal position. After 60 seconds of water flow, the stairwell door was opened. One minute later, the west double door on the north side of the second floor was opened, and water continued to flow for 60 seconds. Next, the water flow was stopped, the two doors were closed, and the procedure was repeated three additional times for the sweeping (back and forth, left-to-right), clockwise, and counterclockwise application patterns. This entire process was repeated using narrow fog and wide fog streams. Fig.~\ref{fig:test_6_pic} contains an image of a straight stream pattern being applied during Test~6 with the flow path fully established.

@@

\section{Handline Experiments}
\label{sec:handline_results}
Fig.~\ref{fig:Test_6_BDP_A10_Avg_All} contains a plot of the average gas velocity measured by the bi-directional probes in the array at the interior stairwell door (A10) for the straight stream, narrow fog stream, and wide fog stream being applied in a fixed position, a sweeping pattern (back and forth, left-to-right), a clockwise rotation, and a counterclockwise rotation during Test~6. Similar plots of the average measured velocity through the interior doorway for Tests~5 and 7 are presented in Appendix~\ref{chap:handline_plots}. Additionally, Tables~\ref{table:east_hand_A6_avgs}~and~\ref{table:west_hand_A10_avgs} list the average velocity measured by the bi-directional probes in A10 when the flow path was fully established for each hose stream and application pattern combination studied during Test~5 and Tests~6--7, respectively.

@@

With the exception of the narrow fog stream in the fixed position during Test~6, every hose stream and application pattern combination using a narrow fog or wide fog stream during the handline experiments produced significant air movement---an average gas velocity greater than 0.5~m/s (1.1~mph)---through the interior doorway while the flow path was fully established. For the straight stream, however, significant air movement through the interior doorway only occurred when the stream was applied in a pattern that involved moving the handline. Similar to the monitor experiments, the average air velocity through the interior doorway was always greater in magnitude for the application patterns using the narrow and wide fog streams than the same patterns using the straight stream. Another important result seen in Tables~\ref{table:east_hand_A6_avgs} and \ref{table:west_hand_A10_avgs} is that there are no statistically significant differences in the amount of air flow through the interior doorway when rotating the hoseline in the clockwise direction compared to rotating it in the counterclockwise direction for any of three hose streams tested. The similarities in air flow produced by the clockwise and counterclockwise application patterns can also be seen graphically in Fig.~\ref{fig:Test_6_BDP_A10_Avg_All} and Figs.~\ref{fig:Test_5_BDP_A6_Avg_All} and \ref{fig:Test_7_BDP_A10_Avg_All} in Appendix~\ref{chap:handline_plots}.

@@

A total of seven experiments were conducted to study the impact of different hose stream and application pattern combinations on air movement within residential scale structures.

Three experiments studied the differences between a straight stream, narrow fog stream, and wide fog stream emitted from a combination nozzle attached to a monitor aimed at specific locations in a structure for different ventilation configurations. For each of the three experiments, the average air velocity through the interior doorway while the flow path was fully established was greatest for the wide fog stream no matter the location and ventilation pattern within the structure. The magnitude of the average air velocity through the interior doorway while the flow path was fully established for the narrow fog stream was always equal to or greater than the magnitude of the average air velocity for the straight stream. Additionally, a fourth experiment using a monitor was conducted in which a straight stream from a combination nozzle was compared to a solid stream from a smooth bore nozzle with a 1~in tip aimed at a specified location for 60 seconds with a fully established flow path. The streams were compared for six different 60 seconds periods during which the flow path was fully established: three times each for the monitor aimed at two locations. The results from the four monitor experiments suggest a wide fog stream impacts air movement within a structure the most out of any of the tested streams, followed by the narrow fog stream, and then the straight stream and solid stream from a smooth bore nozzle, which both caused approximately the same amount of air movement through the structure.

In addition to the monitor experiments, three different experiments were conducted using a handline to study the impact of different stream and application pattern combinations on air movement within a structure. Similar to the monitor experiments, the average air velocity through the interior doorway of a structure with a fully established flow path was greater for the wide fog and narrow fog streams than the straight stream. In fact, it was discovered that a straight stream only caused significant air movement through the interior doorway when it was applied in a moving pattern. Additionally, throughout all handline experiments, there were no statistically significant differences in the average air velocity measured through the interior doorway between the clockwise and counterclockwise application patterns.

Overall, the water flow experiments discussed in this report reveal that hose streams can affect air movement within a structure. The type of hose stream used and manner in which it is applied dictates the extent to which a stream impacts the ventilation of a structure. These experiments have shown there is a potential for hose streams to affect air flows within a structure during fire scenarios. To better quantify the impact of air flows from hose streams on the fire environment, additional fire experiments need to be conducted.

>>>>

%http://www.tft.com/literature/library/files/bfa.pdf
%Keith Royer, "Water for Fire Fighting", (Engineering Extension Service, Iowa State University, Ames, Iowa) p. 6 
In addition to the type of stream pattern used, the method or motion of the nozzle used to apply the hose stream can also vary. For example, Royer suggested using a clockwise rotation instead of a counterclockwise nozzle movement pattern for the following reasons~\cite{Royer:ISU}:
\begin{enumerate} 
	\item Clockwise rotation drives most of the heated gases, smoke, and flames away from the nozzle, while counterclockwise rotation does just the opposite.
	\item Steam formed by the clockwise rotation has a violent rolling action. Counterclockwise rotation produces steam with an inactive and lazy action.
	\item A clockwise rotation increases the efficiency of water and produces a faster knockdown time than counterclockwise rotation.
\end{enumerate}

\section{Objectives}
\label{sec:objectives}
This report is concerned with examining the impact of hose stream and application pattern, or nozzle motion, selection on air movement inside two full scale test structures built to represent residential sized structures. Numerous experiments containing different flow path configurations were conducted to study the impact of hose stream and nozzle movement selection on ventilation in a structure. More specifically, the experiments were conducted to fulfill the following objectives:
\begin{enumerate}
	\item Determine which hose stream patterns and nozzle movement patterns cause significant air movement in a structure 
	\item Determine the amount of air movement induced by different hose streams and nozzle movement patterns

\end{enumerate}
Four types of hose streams and four different nozzle movement patterns were studied throughout the experiments described in this report. Three types of hose stream patterns using a combination nozzle were studied: straight stream, narrow fog stream, and wide fog stream. Water was flowed into the structure by having the hose in a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Additionally, a set of experiments comparing the impact of a straight stream from a combination nozzle and a solid stream from a smooth bore nozzle equipped with a 1~in tip on air flows in a structure was conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

@@

% =========================
% = EXPERIMENTAL OVERVIEW =
% =========================
\chapter{Experimental Overview}
\label{chap:exp_overview}
To study the impact of different hose stream and nozzle movement selections on air movement inside a structure, water flow experiments were conducted in two experimental structures designed to replicate typical residential structures. The experiments focused on studying the differences in the amount of air movement caused by specific combinations of four different hose stream patterns and four different nozzle movement patterns.

@@

A total of seven experiments were conducted to study the impact of different hose stream and nozzle movement patterns on gas movement in a structure; three used a combination nozzle attached to a monitor to flow water at approximately 120~gpm, three used a combination nozzle attached directly to a 1.75~in handline to flow water at approximately 120~gpm (Fig.~\ref{fig:monitor+handline}), and one used both a combination nozzle and a smooth bore nozzle attached to a monitor to flow water at approximately 180~gpm. Experiments that used a monitor to flow water always applied the stream in a fixed position and were focused on changing the location of water application in addition to the stream pattern. The experiments that used a combination nozzle attached directly to a 1.75~in handline to flow water focused on impact of the motion of the nozzle.

@@

Three different structure flow path configurations were used for the seven experiments. Tests~1 and 5 were conducted in the Single Story Structure and used the configuration shown in Fig.~\ref{fig:east_setup}. Test~1 used a monitor and combination nozzle to flow water, and Test~5 used a combination nozzle attached directly to a handline to flow water.

Two different flow path configurations were used for experiments conducted in the Two Story Structure. The main difference between the two flow paths was the position of the south side door on the first level. Tests~2 and 6 used a configuration with the door opened (Fig.~\ref{fig:flow_path_1}), while Tests~3, 4, and 7 used a configuration with the door closed (Fig.~\ref{fig:flow_path_2}). Tests~2 and 3 used a combination nozzle attached to a monitor to flow water; Tests~6 and 7 used a combination nozzle attached directly to a handline to flow water; and Test~4 used both a combination nozzle and a 1~in tip smooth bore nozzle attached to a monitor to flow water in a straight and solid stream, respectively.

@@

\subsubsection{Test~1}
Test~1 utilized a monitor equipped with a combination nozzle to flow water from Room~C of the Single Story Structure to the area above the doorway on the south wall of Room~B (Fig.~\ref{fig:test_1_pic}). A plan view of the structure's layout during Test~1 can be found in Fig.~\ref{fig:east_setup}. The figure highlights the general direction of the established flow path along with the approximate location of A6, the bi-directional probe array used to measure gas velocity through the northeast interior doorway. The procedure for Test~1 was the simplest of all the test procedures. First, water was flowed in a straight stream for 30 seconds, then the stream was changed to a narrow fog for 30 seconds, and finally the stream was changed to a wide fog for 30 seconds. The procedure was repeated.

@@

\subsubsection{Tests~2 \& 3}
Tests~2 and 3 followed identical procedures that involved using a monitor equipped with a combination nozzle to flow water from the north side double doors to the interior ceiling on the first floor of the Two Story Structure. The nozzle was aimed at two regions marked on the center of the ceiling during the series: a ``near'' target located approximately 1.8~m (6~ft) from the interior side of the north wall and a ``far'' target located approximately 3.7~m (12~ft) from the interior side of the north wall. The procedure for each test began by flowing water in a straight stream pattern at the far target for 60 seconds. After 60 seconds of water flow, the stairwell door was opened. One minute later, the west double door on the north side of the second floor was opened, and water continued to flow for 60 seconds. Next, the water flow was stopped, the two doors were closed, and the procedure was repeated with the monitor aimed at the near target. This entire process was repeated using the narrow fog and wide fog streams. Fig.~\ref{fig:test_2_3_pic} contains images of each type of hose stream aimed at the near and far targets.

@@

\subsection{Handline Experiments}
\label{sec:handline_procedure}
Three experiments were conducted using a 1.75~in handline with a combination nozzle to flow water in a straight stream, narrow fog stream, and wide fog stream. One test, Test~5, occurred in the Single Story Structure, and the remaining tests, Tests~6 and 7, occurred in the Two Story Structure. All three tests primarily focused on the impact of different hose stream and nozzle motion pattern selections on air movement in a structure.

@@

\subsubsection{Tests~6 \& 7}
Tests~6 and 7 followed identical procedures that involved using a 1.75~in handline equipped with a combination nozzle to flow water from the north side double doors of the Two Story Structure into the first floor room. The water stream was aimed at the south side wall of the first floor during the experiments. Test~6 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_1}, and Test~7 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_2}. Each experiment began by using a straight stream to flow water in a fixed, horizontal position. After 60 seconds of water flow, the stairwell door was opened. One minute later, the west double door on the north side of the second floor was opened, and water continued to flow for 60 seconds. Next, the water flow was stopped, the two doors were closed, and the procedure was repeated three additional times for the sweeping (back and forth, left-to-right), clockwise, and counterclockwise nozzle motion patterns. This entire process was repeated using narrow fog and wide fog streams. Fig.~\ref{fig:test_6_pic} contains an image of a straight stream pattern being applied during Test~6 with the flow path fully established.

@@

\section{Handline Experiments}
\label{sec:handline_results}
Fig.~\ref{fig:Test_6_BDP_A10_Avg_All} contains a plot of the average gas velocity measured by the bi-directional probes in the array at the interior stairwell door (A10) for the straight stream, narrow fog stream, and wide fog stream being applied in a fixed position, a sweeping pattern (back and forth, left-to-right), a clockwise rotation, and a counterclockwise rotation during Test~6. Similar plots of the average measured velocity through the interior doorway for Tests~5 and 7 are presented in Appendix~\ref{chap:handline_plots}. Additionally, Tables~\ref{table:east_hand_A6_avgs}~and~\ref{table:west_hand_A10_avgs} list the average velocity measured by the bi-directional probes in A10 when the flow path was fully established for each hose stream and nozzle movement pattern combination studied during Test~5 and Tests~6--7, respectively.

@@

With the exception of the narrow fog stream in the fixed position during Test~6, every hose stream and nozzle movement pattern combination using a narrow fog or wide fog stream during the handline experiments produced significant air movement---an average gas velocity greater than 0.5~m/s (1.1~mph)---through the interior doorway while the flow path was fully established. For the straight stream, however, significant air movement through the interior doorway only occurred when the stream was applied in a pattern that involved moving the handline. Similar to the monitor experiments, the average air velocity through the interior doorway was always greater in magnitude for the nozzle movement patterns using the narrow and wide fog streams than the same patterns using the straight stream. Another important result seen in Tables~\ref{table:east_hand_A6_avgs} and \ref{table:west_hand_A10_avgs} is that there are no statistically significant differences in the amount of air flow through the interior doorway when rotating the hoseline in the clockwise direction compared to rotating it in the counterclockwise direction for any of three hose streams tested. The similarities in air flow produced by the clockwise and counterclockwise nozzle movement patterns can also be seen graphically in Fig.~\ref{fig:Test_6_BDP_A10_Avg_All} and Figs.~\ref{fig:Test_5_BDP_A6_Avg_All} and \ref{fig:Test_7_BDP_A10_Avg_All} in Appendix~\ref{chap:handline_plots}.

@@

A total of seven experiments were conducted to study the impact of different hose stream and nozzle movement pattern combinations on air movement within residential scale structures.

Three experiments studied the differences between a straight stream, narrow fog stream, and wide fog stream emitted from a combination nozzle attached to a monitor aimed at specific locations in a structure for different ventilation configurations. For each of the three experiments, the average air velocity through the interior doorway while the flow path was fully established was greatest for the wide fog stream no matter the location and ventilation pattern within the structure. The magnitude of the average air velocity through the interior doorway while the flow path was fully established for the narrow fog stream was always equal to or greater than the magnitude of the average air velocity for the straight stream.

Additionally, a fourth experiment using a monitor was conducted in which a straight stream from a combination nozzle was compared to a solid stream from a smooth bore nozzle with a 1~in tip aimed at a specified location for 60 seconds with a fully established flow path. The streams were compared for six different 60 seconds periods during which the flow path was fully established: three times each for the monitor aimed at two locations. The results from the four monitor experiments suggest a wide fog stream impacts air movement within a structure the most out of any of the tested streams, followed by the narrow fog stream, and then the straight stream and solid stream from a smooth bore nozzle, which both caused approximately the same amount of air movement through the structure.

In addition to the monitor experiments, three different experiments were conducted using a handline to study the impact of different stream and nozzle movement pattern combinations on air movement within a structure. Similar to the monitor experiments, the average air velocity through the interior doorway of a structure with a fully established flow path was greater for the wide fog and narrow fog streams than the straight stream. In fact, it was discovered that a straight stream only caused significant air movement through the interior doorway when it was applied in a moving pattern. Additionally, throughout all handline experiments, there were no statistically significant differences in the average air velocity measured through the interior doorway between the clockwise and counterclockwise nozzle movement patterns.

Overall, the water flow experiments discussed in this report reveal that hose streams can affect air movement within a structure. The type of hose stream used and manner in which it is applied dictates the extent to which a stream impacts the ventilation of a structure. These experiments have shown there is a potential for hose streams to affect air flows within a structure during fire scenarios. To better quantify the impact of air flows from hose streams on the fire environment, additional fire experiments need to be conducted with structure fires in a controlled environment.
